% ==============================================================================
% Section 7: Mathematical Instantiation
% "Endogenous Standpoint as a Low-Curvature Attractor"
% ==============================================================================

\section{Mathematical Instantiation: Transformer Attention as Discrete Vector Bundle Connection}
\label{sec:instantiation}

This section establishes the central mathematical result of the paper: a transformer's
attention mechanism defines a discrete connection on a vector bundle over the event-sequence
space, and its holonomy deviation \emph{is} the path-dependence that measures standpoint
drift~\cite{LCESA} and binding failure~\cite{self-jurisdiction}. Unlike classical parallel
transport, this connection is not necessarily metric-compatible; the holonomy deviation
measures the \emph{differential} transport across standpoint layers, which is the
geometrically meaningful quantity. We proceed by instantiating each component of the
abstract fiber bundle framework (Definitions~1--8 of~\cite{LCESA}, reproduced here for
self-containedness) in terms of concrete transformer architectural objects: residual
streams, attention heads, value matrices, and attention weights.

Throughout, we fix the following notation. Let $d$ denote the residual stream dimension,
$H$ the number of attention heads, $d_v$ the per-head value dimension, and $L$ the number
of attention layers. For standpoint layer $k \in K = \{\mathrm{min}, \mathrm{nar},
\mathrm{soc}, \mathrm{mor}, \mathrm{pos}\}$, we write $d_k = \dim(W_k)$ for the dimension
of the corresponding subspace, so that $d = \sum_{k \in K} d_k$.

% ------------------------------------------------------------------------------
\subsection{Event Space from Conversation Structure}
\label{sec:event-space}

We begin by instantiating the base space of the fiber bundle. Following Definition~1
of~\cite{LCESA}, the event space captures the causal and temporal structure of a
multi-turn conversation.

\begin{definition}[Conversation Event Space, {\cite[Def.~1]{LCESA}}]
\label{def:event-space}
Let $\mathcal{X} = (V, E, T)$ be a simplicial complex defined as follows:
\begin{enumerate}
    \item \textbf{Vertices (turns).} $V = \{x_1, \ldots, x_n\}$, where each $x_i$ is a
    conversation turn---a contiguous segment of tokens produced by a single interlocutor
    in a multi-turn dialogue.
    \item \textbf{Edges (causal order).} $E = \{(x_i, x_j) : j = i+1 \text{ or } x_j
    \text{ causally depends on } x_i\}$, a set of directed edges following turn order and
    causal dependency.
    \item \textbf{2-simplices (consistency obligations).} $T = \{\langle x_i, x_j, x_k
    \rangle \in V^3 : \text{these three turns impose a consistency obligation}\}$. For
    example, $x_i$ is an assertion, $x_j$ is a pressure turn (e.g., a challenge or
    suggestion to revise), and $x_k$ is a subsequent output that should cohere with $x_i$.
\end{enumerate}
\end{definition}

\begin{remark}[Turn-level granularity]
\label{rem:turn-granularity}
We take the vertex set $V$ at the level of conversation \emph{turns}, not individual
tokens. This choice is deliberate and reflects the semantics of the standpoint: a
standpoint is a commitment that is maintained (or fails to be maintained) across
\emph{propositional units}, not across individual sub-word tokens. The token-level
sequence is the domain of the transformer's internal computation, but the event space
over which the fiber bundle is defined lives at the semantic grain of turns. The mapping
from token-level hidden states to turn-level events is given by taking the residual
stream at the final token of each turn (see \S\ref{sec:fiber}).
\end{remark}

% ------------------------------------------------------------------------------
\subsection{Fiber from Head Value Subspaces}
\label{sec:fiber}

We now instantiate the fiber of the bundle. The key idea is that the transformer's
residual stream decomposes into orthogonal subspaces, each associated with a standpoint
layer and spanned by the value matrices of a designated group of attention heads.

\begin{definition}[Head Grouping Function, {\cite[Def.~2]{LCESA}}]
\label{def:head-grouping}
Let $\gamma : \{1, \ldots, H\} \to K$ be a function assigning each attention head to
exactly one standpoint layer. For layer $k \in K$, define the head set
\[
    H_k = \{h \in \{1, \ldots, H\} : \gamma(h) = k\}.
\]
Each head $h$ has an associated value matrix $V_h \in \mathbb{R}^{d \times d_v}$. The
\textbf{standpoint subspace} for layer $k$ is
\[
    W_k = \operatorname{span}\!\Big(\bigcup_{h \in H_k} \operatorname{range}(V_h)\Big)
    \subseteq \mathbb{R}^d.
\]
\end{definition}

\begin{assumption}[Subspace Independence]
\label{ass:independence}
The standpoint subspaces $\{W_k\}_{k \in K}$ are \textbf{independent}: the sum
$\sum_k W_k$ is direct (i.e., $W_k \cap \operatorname{span}(\bigcup_{l \neq k} W_l)
= \{0\}$ for all $k$), and $\sum_{k \in K} d_k = d$, so that
$\mathbb{R}^d = \bigoplus_k W_k$.
\end{assumption}

\begin{remark}[Orthogonality as an Empirical Condition]
\label{rem:orthogonality}
Assumption~\ref{ass:independence} does \emph{not} require the subspaces to be mutually
orthogonal. Orthogonality ($W_k \perp W_l$ for $k \neq l$) is a sufficient condition for
independence, but not a necessary one. In practice, the degree of orthogonality is an
empirical quantity that can be measured via the maximum principal angle:
\[
    \delta = \max_{k \neq l}\, \sin\theta_{\max}(W_k, W_l)
    = \max_{k \neq l}\, \|P_{W_k} P_{W_l}\|_2.
\]
When $\delta = 0$, the subspaces are orthogonal and the classical framework applies
exactly. When $\delta > 0$ but small, the oblique projector framework
(Definition~\ref{def:oblique-projector}) provides an exact treatment, and all results
hold with the appropriate projectors. The quantity $\delta$ is experimentally measurable
from the trained model's value matrices.
\end{remark}

\begin{definition}[Oblique Projector]
\label{def:oblique-projector}
For the independent subspaces $\{W_k\}_{k \in K}$ of Assumption~\ref{ass:independence},
define the \textbf{oblique projector} $\Pi_k : \mathbb{R}^d \to W_k$ by
\[
    \Pi_k = W_k\, (Z_k^\top W_k)^{-1}\, Z_k^\top,
\]
where $W_k \in \mathbb{R}^{d \times d_k}$ is a basis matrix for $W_k$ and $Z_k \in
\mathbb{R}^{d \times d_k}$ is a basis matrix for $(\operatorname{span}(\bigcup_{l \neq k}
W_l))^\perp$. The oblique projector satisfies:
\begin{enumerate}
    \item $\Pi_k^2 = \Pi_k$ (idempotent),
    \item $\operatorname{range}(\Pi_k) = W_k$,
    \item $\Pi_k\, \Pi_l = 0$ for $k \neq l$ (mutually annihilating),
    \item $\sum_{k \in K} \Pi_k = I_d$ (resolution of identity).
\end{enumerate}
When the subspaces are orthogonal, $\Pi_k = P_{W_k}$ (the orthogonal projector), so this
framework subsumes the orthogonal case.
\end{definition}

\begin{proposition}[Fiber Decomposition]
\label{prop:fiber-decomposition}
Under Assumption~\ref{ass:independence}, the residual stream admits the direct sum
decomposition
\[
    \mathbb{R}^d = W_{\mathrm{min}} \oplus W_{\mathrm{nar}} \oplus W_{\mathrm{soc}}
    \oplus W_{\mathrm{mor}} \oplus W_{\mathrm{pos}},
\]
with $\sum_{k \in K} d_k = d$, where $d_k = \dim(W_k) = |H_k| \cdot d_v$. The fiber at
event $x_i$ is $E_{x_i} = \mathbb{R}^d$ equipped with this decomposition and the oblique
projectors $\{\Pi_k\}_{k \in K}$ of Definition~\ref{def:oblique-projector}.
\end{proposition}

\begin{proof}
By Assumption~\ref{ass:independence}, the subspaces are independent and
$\sum_k d_k = d$, so the sum is direct and fills $\mathbb{R}^d$. By construction,
every attention head $h \in \{1,\ldots,H\}$ belongs to exactly one $H_k$, and
$\bigcup_h \operatorname{range}(V_h)$ spans $\mathbb{R}^d$ (the output projection matrix
$W_O$ is full rank in standard transformer architectures). Therefore $\bigoplus_k W_k =
\mathbb{R}^d$. The resolution of identity $\sum_k \Pi_k = I_d$ follows from
Definition~\ref{def:oblique-projector}.
\end{proof}

\begin{definition}[Standpoint State]
\label{def:standpoint-state}
The \textbf{standpoint state} at event $x_i$ is the tuple
\[
    \Psi_{x_i} = \big(\psi_{\mathrm{min}}(x_i),\; \psi_{\mathrm{nar}}(x_i),\;
    \psi_{\mathrm{soc}}(x_i),\; \psi_{\mathrm{mor}}(x_i),\; \psi_{\mathrm{pos}}(x_i)\big)
    \in \bigoplus_{k \in K} W_k,
\]
where $\psi_k(x_i) = \Pi_k\, h_{x_i}$ is the oblique projection of the residual stream
$h_{x_i} \in \mathbb{R}^d$ at the final token of turn $x_i$ onto the subspace $W_k$.
\end{definition}

\begin{definition}[Structure Group, {\cite[Def.~3]{LCESA}}]
\label{def:structure-group}
The \textbf{structure group} of the fiber bundle is
\[
    G = O(W_{\mathrm{min}}) \times O(W_{\mathrm{nar}}) \times O(W_{\mathrm{soc}})
    \times O(W_{\mathrm{mor}}) \times O(W_{\mathrm{pos}})
    \cong O(d_{\mathrm{min}}) \times O(d_{\mathrm{nar}}) \times O(d_{\mathrm{soc}})
    \times O(d_{\mathrm{mor}}) \times O(d_{\mathrm{pos}}).
\]
An element $g = (g_{\mathrm{min}}, g_{\mathrm{nar}}, g_{\mathrm{soc}}, g_{\mathrm{mor}},
g_{\mathrm{pos}}) \in G$ acts on the fiber $\mathbb{R}^d$ by
\[
    g \cdot h = \sum_{k \in K} g_k\, \Pi_k\, h,
\]
applying an independent orthogonal transformation within each standpoint subspace. A
\textbf{gauge transformation} of the transport operators is
\[
    U_{ij} \mapsto g_j\, U_{ij}\, g_i^{-1}, \qquad g_i, g_j \in G.
\]
The fiber metric $g_F$ is $G$-invariant, and the structure group $G$ defines the gauge
freedom: two connection representations related by $U_{ij} \mapsto g_j U_{ij} g_i^{-1}$
(with $g_i, g_j \in G$) are gauge-equivalent. This gauge symmetry is the basis for the
gauge-invariant diagnostic framework developed in Section~\ref{sec:fnsi}. We note that the
actual transport operators constructed from transformer attention
(Section~\ref{sec:transport}) are scalar scalings on each block, lying in
$\mathbb{R}_{>0} \subset GL(d_k, \mathbb{R})$ rather than $O(d_k)$; the framework is
accordingly a vector bundle connection with $G$-gauge symmetry, as clarified in
Remark~\ref{rem:vector-bundle}.
\end{definition}

\begin{remark}[Framework Clarification: Vector Bundle Connection]
\label{rem:vector-bundle}
The transport operators $U_{ij}$ constructed in Section~\ref{sec:transport} lie in
$GL(d_k, \mathbb{R})$ on each standpoint block, not in $O(d_k)$. Specifically, under the
block-diagonal structure (Proposition~\ref{prop:block-structure}), the transport on block
$W_k$ is $\bar{\alpha}^{(k)}_{ji} \cdot \mathrm{Id}_{d_k}$, which is a scalar scaling map
and generally not an orthogonal transformation unless $\bar{\alpha}^{(k)}_{ji} = 1$. The
present framework is therefore best understood as a \textbf{vector bundle with a linear
connection}, rather than a principal bundle with a $G$-connection. The structure group
$G = \prod_k O(d_k)$ defines the gauge symmetry of basis choices within each standpoint
subspace --- the freedom to rotate the orthonormal basis of $W_k$ without changing the
holonomy norms --- but $G$ is not the holonomy group of the connection. All results in
this paper, including Gauge Invariance (Propositions~\ref{prop:gauge-abstract},
\ref{prop:gauge-invariance}) and the Block Decomposition (Lemma~\ref{lem:decomposition},
Proposition~\ref{prop:block-structure}), hold for this more general vector-bundle setting.
The gauge invariance proofs are unaffected: they show that $\|[F_{ijk}]_k - \Pi_k\|_F$ is
invariant under $G$-conjugation, which is a property of orthogonal basis transformations
within $W_k$, not of the full holonomy group.
\end{remark}

\begin{definition}[Canonical Fiber Metric, {\cite[Def.~4]{LCESA}}]
\label{def:fiber-metric}
The \textbf{canonical fiber metric} on $E_{x_i} = \mathbb{R}^d$ is
\[
    g_F = \alpha_{\mathrm{min}}\, I_{d_{\mathrm{min}}} \oplus
    \alpha_{\mathrm{nar}}\, I_{d_{\mathrm{nar}}} \oplus
    \alpha_{\mathrm{soc}}\, I_{d_{\mathrm{soc}}} \oplus
    \alpha_{\mathrm{mor}}\, I_{d_{\mathrm{mor}}} \oplus
    \alpha_{\mathrm{pos}}\, I_{d_{\mathrm{pos}}},
\]
where $\alpha_k > 0$ are layer weights. For the present instantiation we take $\alpha_k
= 1$ for all $k$ (the standard Euclidean metric). The resulting inner product is
$\langle h, h' \rangle_F = \sum_{k \in K} \langle \Pi_k h,\, \Pi_k h' \rangle$, where
$\Pi_k$ are the oblique projectors of Definition~\ref{def:oblique-projector}. When the
subspaces are orthogonal, $\Pi_k = P_{W_k}$ and this reduces to the standard Euclidean
inner product.
\end{definition}

% ------------------------------------------------------------------------------
\subsection{Transport from Attention}
\label{sec:transport}

We now construct the connection on the fiber bundle. A connection is a rule for
transporting fiber elements along the base space: given the standpoint state at event
$x_i$, how should it be carried to event $x_j$? We show that the transformer's attention
mechanism defines such a connection. Unlike the classical setting of parallel transport on
Riemannian manifolds, this connection is not necessarily metric-compatible
(Remark~\ref{rem:metric-preservation}): attention can differentially amplify or attenuate
different standpoint layers. The holonomy deviation of this connection, defined in
\S\ref{sec:curvature}, measures the inconsistency of transport across layers.

\begin{definition}[Attention Transport Operator, {\cite[Def.~6]{LCESA}}]
\label{def:transport-operator}
The \textbf{transport operator} from event $x_i$ to event $x_j$ is the linear map
$U_{ij} : \mathbb{R}^d \to \mathbb{R}^d$ defined by
\[
    U_{ij}(h) = \sum_{h=1}^{H} \alpha_{ji}^{(h)}\, V_h\, V_h^\top\, h,
\]
where:
\begin{itemize}
    \item $\alpha_{ji}^{(h)} \in [0,1]$ is the attention weight of head $h$ from the
    last token of event $x_i$ to event $x_j$, aggregated over all tokens in $x_j$.
    \item $V_h \in \mathbb{R}^{d \times d_v}$ is the value projection matrix of head $h$.
    \item $V_h V_h^\top \in \mathbb{R}^{d \times d}$ is the orthogonal projection onto
    $\operatorname{range}(V_h)$ (assuming $V_h$ has orthonormal columns).
\end{itemize}
\end{definition}

\begin{remark}[Orthonormality Assumption]
\label{rem:orthonormality}
The derivation of $V_h V_h^\top = P_{\mathrm{range}(V_h)}$ requires $V_h^\top V_h =
\mathrm{Id}_{d_v}$, i.e., the columns of $V_h$ are orthonormal. This is not enforced by
transformer training. We assess this assumption empirically by computing, for each head
$h$, the deviation $\epsilon_h = \|V_h^\top V_h - \mathrm{Id}_{d_v}\|_F$. The mean and
distribution of $\epsilon_h$ across all heads are reported in Appendix~\ref{app:orthonorm}.
When $\epsilon_h$ is small (columns are approximately orthonormal), the derived
block-diagonal structure holds approximately. When $\epsilon_h$ is large, the transport
operator $U_{ij}$ acquires off-diagonal cross-block contributions beyond those captured by
the non-orthogonality parameter $\delta$. All holonomy computations use the empirically
corrected projectors described in Appendix~\ref{app:orthonorm}.
\end{remark}

\begin{proposition}[Block Structure of $U_{ij}$]
\label{prop:block-structure}
Under Assumption~\ref{ass:independence} and the assumption that $\{V_h : h \in H_k\}$
form an orthonormal set spanning $W_k$ for each $k$, the transport operator satisfies
\[
    U_{ij} = \sum_{k \in K} \bar{\alpha}_{ji}^{(k)}\, P_{W_k} + E_{ij},
\]
where $\bar{\alpha}_{ji}^{(k)} = \frac{1}{|H_k|} \sum_{h \in H_k} \alpha_{ji}^{(h)}$ is
the mean attention weight of layer $k$'s heads from $x_i$ to $x_j$, $P_{W_k}$ is the
orthogonal projector onto $W_k$, and $E_{ij}$ is the cross-layer error term with
\[
    \|E_{ij}\|_F \leq \delta \cdot \sum_{k \in K} |H_k| \cdot \max_{h \in H_k}
    |\alpha_{ji}^{(h)}|,
\]
where $\delta = \max_{k \neq l} \|P_{W_k} P_{W_l}\|_2$ measures subspace non-orthogonality
(Remark~\ref{rem:orthogonality}).

When the subspaces are orthogonal ($\delta = 0$), $E_{ij} = 0$ and the transport operator
is exactly block-diagonal:
\[
    U_{ij} = \operatorname{diag}\!\big(\bar{\alpha}_{ji}^{(\mathrm{min})}\,
    I_{d_{\mathrm{min}}},\; \bar{\alpha}_{ji}^{(\mathrm{nar})}\, I_{d_{\mathrm{nar}}},\;
    \ldots,\; \bar{\alpha}_{ji}^{(\mathrm{pos})}\, I_{d_{\mathrm{pos}}}\big).
\]
\end{proposition}

\begin{proof}
By the head grouping function $\gamma$, each head $h \in H_k$ has
$\operatorname{range}(V_h) \subseteq W_k$, and the sets $\{V_h\}_{h \in H_k}$ are
orthonormal and span $W_k$. Therefore $\sum_{h \in H_k} V_h V_h^\top = P_{W_k}$, the
orthogonal projector onto $W_k$. Summing over heads:
\[
    U_{ij} = \sum_{h=1}^{H} \alpha_{ji}^{(h)} V_h V_h^\top
           = \sum_{k \in K} \sum_{h \in H_k} \alpha_{ji}^{(h)} V_h V_h^\top
           = \sum_{k \in K} \bar{\alpha}_{ji}^{(k)} P_{W_k}.
\]
This is the \emph{exact} expression for $U_{ij}$; no approximation has been made. The
block structure depends on the orthogonality of the $\{P_{W_k}\}$. Decomposing $U_{ij}$
in the block basis $\mathbb{R}^d = \bigoplus_k W_k$:
\[
    [U_{ij}]_{kl} = P_{W_k}\, U_{ij}\, P_{W_l}
    = \bar{\alpha}_{ji}^{(k)}\, P_{W_k}\, P_{W_l}.
\]
When $\delta = 0$ (orthogonal subspaces), $P_{W_k} P_{W_l} = \delta_{kl} P_{W_k}$ and
$U_{ij}$ is block-diagonal. When $\delta > 0$, the off-diagonal blocks satisfy
$\|[U_{ij}]_{kl}\|_F \leq |\bar{\alpha}_{ji}^{(k)}| \cdot \delta \cdot \sqrt{d_k \cdot
d_l}$ for $k \neq l$, and the stated bound on $E_{ij}$ follows.
\end{proof}

\begin{remark}[Metric Compatibility and Connection Type]
\label{rem:metric-preservation}
The transport operator $U_{ij}$ is \emph{not} in general an isometry of the fiber metric
$g_F$. We distinguish three regimes:
\begin{enumerate}
    \item \textbf{Isometric transport} ($\bar{\alpha}_{ji}^{(k)} = 1$ for all $k$):
    $\langle U_{ij} h, U_{ij} h' \rangle_F = \langle h, h' \rangle_F$. This is strict
    parallel transport.
    \item \textbf{Conformal transport} ($\bar{\alpha}_{ji}^{(k)} = \lambda$ for all $k$,
    some $\lambda > 0$): $\langle U_{ij} h, U_{ij} h' \rangle_F = \lambda^2 \langle h, h'
    \rangle_F$. All layers are scaled equally.
    \item \textbf{Non-metric transport} (general case): Different layers are scaled by
    different amounts. This is the generic situation in a transformer.
\end{enumerate}
In all three cases, $U_{ij}$ defines a \emph{discrete connection} on the fiber bundle
(not necessarily metric-compatible). The curvature (Definition~\ref{def:curvature})
measures the \emph{differential} transport across layers --- the inconsistency between
how different standpoint layers are transported --- which is well-defined for any
connection, regardless of metric compatibility.
\end{remark}

\begin{proposition}[Holonomy Deviation as Inter-Layer Inconsistency]
\label{prop:curvature-inconsistency}
Under the block-diagonal structure ($\delta = 0$), the block-specific holonomy norm has
the explicit form
\[
    \|F_{ijk}\|_k = \left|\frac{\bar{\alpha}_{ji}^{(k)}\, \bar{\alpha}_{kj}^{(k)}}
    {\bar{\alpha}^{(k)}_{\exp}} - 1\right| \cdot \sqrt{d_k}.
\]
The holonomy deviation is zero if and only if the composite attention weight for layer $k$ matches
the baseline expectation. The holonomy deviation \emph{across} layers measures the inconsistency of
transport rates: when $\|F_{ijk}\|_{\mathrm{nar}} \neq \|F_{ijk}\|_{\mathrm{mor}}$, the
narrative and moral layers are being transported at different rates relative to their
baselines, which is the geometric signature of standpoint drift.
\end{proposition}

% ------------------------------------------------------------------------------
\subsection{Holonomy Deviation as Path-Dependence}
\label{sec:curvature}

The holonomy deviation of a connection on a fiber bundle measures the failure of parallel
transport to be path-independent. We now instantiate this classical notion in the
transformer setting, where it acquires a concrete operational meaning: holonomy deviation
is the path-dependence of attention transport, and it quantifies standpoint drift.

\begin{definition}[Holonomy Deviation, {\cite[Def.~8]{LCESA}}]
\label{def:curvature}
For each 2-simplex $\sigma = \langle x_i, x_j, x_k \rangle \in T$ (a consistency
obligation in the event space), define the \textbf{holonomy deviation} at $\sigma$ by
\[
    F_{ijk} = U_{ij} \cdot U_{jk} \cdot \big(U_{ik}^{\exp}\big)^{-1} \in
    \operatorname{GL}(d, \mathbb{R}),
\]
where $U_{ik}^{\exp}$ is the \emph{expected flat transport} from $x_i$ to $x_k$, defined
as the transport that would occur if the intermediate event $x_j$ had no effect on the
standpoint. Concretely, $U_{ik}^{\exp}$ is estimated from baseline (Type~1) conversations
in which the standpoint is preserved:
\[
    U_{ik}^{\exp} = \mathbb{E}_{\text{T1}}\!\big[U_{ij'} \cdot U_{j'k}\big],
\]
where the expectation ranges over baseline conversations in which acknowledged revision
with evidence preserves standpoint coherence.
\end{definition}

\begin{remark}[Practical Estimation]
When a direct estimate of $U_{ik}^{\exp}$ is unavailable, one may approximate
$U_{ik}^{\exp} \approx I_d$ (the identity, representing no transport) and compute
$F_{ijk} \approx U_{ij} \cdot U_{jk}$. The holonomy deviation is then normalized against a
baseline: $F_{ijk}^{\mathrm{norm}} = F_{ijk} \cdot (F_{ijk}^{\mathrm{baseline}})^{-1}$.
\end{remark}

\begin{theorem}[Holonomy Deviation as Path-Dependence]
\label{thm:path-dependence}
$F_{ijk} = I_d$ if and only if the transport from $x_i$ to $x_k$ is the same whether or
not the intermediate event $x_j$ intervenes. That is,
\[
    F_{ijk} = I_d \quad\Longleftrightarrow\quad U_{ij} \cdot U_{jk} = U_{ik}^{\exp}.
\]
\end{theorem}

\begin{proof}
($\Rightarrow$) Suppose $F_{ijk} = I_d$. By Definition~\ref{def:curvature},
\[
    U_{ij} \cdot U_{jk} \cdot (U_{ik}^{\exp})^{-1} = I_d.
\]
Multiplying both sides on the right by $U_{ik}^{\exp}$ gives $U_{ij} \cdot U_{jk} =
U_{ik}^{\exp}$. Thus the composite transport through $x_j$ equals the expected direct
transport: the intermediate event did not alter the standpoint, and transport is
path-independent.

($\Leftarrow$) Suppose $U_{ij} \cdot U_{jk} = U_{ik}^{\exp}$. Then
\[
    F_{ijk} = U_{ij} \cdot U_{jk} \cdot (U_{ik}^{\exp})^{-1}
            = U_{ik}^{\exp} \cdot (U_{ik}^{\exp})^{-1} = I_d.
\]
Hence the holonomy deviation is trivial.
\end{proof}

\begin{definition}[Block-Specific Holonomy Norm]
\label{def:binding-failure}
For standpoint layer $k \in K$, define the \textbf{block-specific holonomy deviation}
\[
    [F_{ijk}]_k = \Pi_k\, F_{ijk}\, \Pi_k \in \operatorname{Mat}(d, \mathbb{R}),
\]
and the \textbf{block-specific holonomy norm}
\[
    \|F_{ijk}\|_k = \big\|[F_{ijk}]_k - \Pi_k\big\|_F.
\]
\end{definition}

\begin{theorem}[Block-Specific Holonomy Norm Detects Binding Failure]
\label{thm:binding-failure}
For standpoint layer $k \in K$ with non-trivial standpoint state ($\psi_k(x_i) \neq 0$):
\[
    \|F_{ijk}\|_k > 0 \quad\Longleftrightarrow\quad \text{the commitment in layer $k$
    was not preserved through event $x_j$ (binding failure in layer $k$)}.
\]
\end{theorem}

\begin{proof}
Under the block-diagonal approximation (Proposition~\ref{prop:block-structure} with
$\delta = 0$), $U_{ij} = \sum_k \bar{\alpha}_{ji}^{(k)} P_{W_k}$ and the holonomy deviation
$F_{ijk}$ is also block-diagonal. The $k$-th block satisfies
$[F_{ijk}]_k = (\bar{\alpha}_{ji}^{(k)} \bar{\alpha}_{kj}^{(k)} /
\bar{\alpha}^{(k)}_{\exp})\, I_{d_k}$, so $\|F_{ijk}\|_k = 0$ if and only if
$\bar{\alpha}_{ji}^{(k)} \bar{\alpha}_{kj}^{(k)} = \bar{\alpha}^{(k)}_{\exp}$.

($\Leftarrow$) If the commitment in layer $k$ was not preserved, then the standpoint
component $\psi_k$ changed non-trivially between $x_i$ and $x_k$ due to the intervening
event $x_j$. This means $\bar{\alpha}_{ji}^{(k)} \bar{\alpha}_{kj}^{(k)} \neq
\bar{\alpha}^{(k)}_{\exp}$, so $[F_{ijk}]_k \neq \Pi_k$ and $\|F_{ijk}\|_k > 0$.

($\Rightarrow$) If $\|F_{ijk}\|_k > 0$, then $[F_{ijk}]_k \neq \Pi_k$, which means the
composite transport through $W_k$ differs from the expected transport. The event $x_j$
introduced an unexpected change in layer $k$'s standpoint component---a commitment that
was not preserved.

For the non-orthogonal case ($\delta > 0$), the block-specific holonomy norm acquires an
$O(\delta)$ correction from cross-layer leakage, but the binding failure characterization
remains valid as long as the curvature signal exceeds the $O(\delta)$ noise floor.
\end{proof}

\begin{remark}[Block Decomposition: Orthogonal vs.\ Non-Orthogonal]
\label{rem:block-decomposition}
When the subspaces are orthogonal ($\delta = 0$), the total curvature decomposes exactly:
\[
    \|F_{ijk} - I_d\|_F^2 = \sum_{k \in K} \big\|[F_{ijk}]_k - I_{d_k}\big\|_F^2
    + \sum_{\substack{k,l \in K \\ k \neq l}} \big\|P_{W_k}\, F_{ijk}\, P_{W_l}\big\|_F^2.
\]
This follows from the Pythagorean theorem for the Frobenius norm, using $P_{W_k} P_{W_l}
= \delta_{kl} P_{W_k}$.

When $\delta > 0$, the oblique projectors $\Pi_k$ satisfy $\Pi_k \Pi_l = 0$ for $k \neq
l$, but the Frobenius inner product between blocks $\Pi_k M \Pi_l$ and $\Pi_{k'} M
\Pi_{l'}$ does not generally vanish for $(k,l) \neq (k',l')$. The total curvature norm
acquires cross-block correction terms:
\[
    \|F_{ijk} - I_d\|_F^2 = \sum_{k,l} \|\Pi_k(F_{ijk} - I_d)\Pi_l\|_F^2
    + O(\delta).
\]
The block-specific norms $\|[F_{ijk}]_k - \Pi_k\|_F$ remain well-defined and
experimentally meaningful as the primary diagnostic quantity.
\end{remark}

% ------------------------------------------------------------------------------
\subsection{Dynamic Non-Commutativity from Layer Composition}
\label{sec:non-abelian}

The block-diagonal structure of $U_{ij}$ (Proposition~\ref{prop:block-structure}) might
suggest that the transport operators commute---that is, that the structure group acts by
independent rotations within each block. We show that this is false at the level of
multi-layer dynamics: the non-commutativity of composed transport arises not from
cross-talk within a single attention layer, but from the \emph{dynamic dependence} of
layer $l+1$'s attention weights on layer $l$'s output.

\begin{definition}[Effective Multi-Layer Transport]
\label{def:effective-transport}
Let $U_{ij}^{(l)}$ denote the transport operator of attention layer $l$ from event $x_i$
to event $x_j$, as in Definition~\ref{def:transport-operator}. The \textbf{effective
transport} through all $L$ attention layers is
\[
    U_{ij}^{\mathrm{eff}} = \prod_{l=1}^{L} \big(I_d + U_{ij}^{(l)}\big),
\]
where the product is ordered from layer $1$ (bottom) to layer $L$ (top), matching the
residual stream accumulation $h_{\mathrm{out}} = h_{\mathrm{in}} + \sum_{l=1}^{L}
\mathrm{attn}^{(l)}(h_{\mathrm{in}})$.
\end{definition}

\begin{proposition}[Cross-Term Coupling]
\label{prop:cross-term}
Consider two attention layers $A$ and $B$ with $U_{ij}^{(A)}$ supported on $W_{\mathrm{nar}}$
and $U_{ij}^{(B)}$ supported on $W_{\mathrm{mor}}$ (i.e., layer $A$'s heads are grouped
to $\mathrm{nar}$ and layer $B$'s heads to $\mathrm{mor}$). Then the effective two-layer
transport is
\[
    U_{ij}^{AB} = \big(I_d + U_{ij}^{B}\big)\big(I_d + U_{ij}^{A}\big)
    = I_d + U_{ij}^{A} + U_{ij}^{B} + U_{ij}^{B}\, U_{ij}^{A}.
\]
The cross-term $U_{ij}^{B}\, U_{ij}^{A}$ maps $W_{\mathrm{nar}} \to W_{\mathrm{mor}}$
(non-trivially), creating inter-layer coupling that does not exist in the single-layer
transport operators. In particular, $U_{ij}^{AB} \neq U_{ij}^{BA}$ in general: the
composed transport is non-commutative (Remark~\ref{rem:non-abelian-source}).
\end{proposition}

\begin{proof}
By hypothesis, $U_{ij}^{(A)} = \bar{\alpha}_{ji}^{A}\, P_{W_{\mathrm{nar}}}$ and
$U_{ij}^{(B)} = \bar{\alpha}_{ji}^{B}\, P_{W_{\mathrm{mor}}}$. Their product is
\[
    U_{ij}^{B}\, U_{ij}^{A}
    = \bar{\alpha}_{ji}^{B}\, \bar{\alpha}_{ji}^{A}\, P_{W_{\mathrm{mor}}}\,
      P_{W_{\mathrm{nar}}}.
\]
By Assumption~\ref{ass:independence}, $W_{\mathrm{nar}}$ and $W_{\mathrm{mor}}$ are
independent but \emph{not necessarily orthogonal}; thus $P_{W_{\mathrm{mor}}}\,
P_{W_{\mathrm{nar}}} \neq 0$ in general. The cross-term $U_{ij}^{B}\, U_{ij}^{A} =
\bar{\alpha}_{ji}^{B}\, \bar{\alpha}_{ji}^{A}\, P_{W_{\mathrm{mor}}}\, P_{W_{\mathrm{nar}}}$
is therefore non-zero as a map from $W_{\mathrm{nar}}$ to $W_{\mathrm{mor}}$.

The deeper point is that in a multi-layer transformer, the residual stream at the input of
layer $B$ is $h + U_{ij}^{(A)}(h)$---layer $B$ sees layer $A$'s output, not the raw
input. Layer $B$'s attention pattern $\alpha_{ji}^{B,(h)}$ is computed from query and key
projections of this modified residual stream. Thus the attention \emph{weights} of layer
$B$ depend on information that layer $A$ placed into $W_{\mathrm{nar}}$. The result is
that $U_{ij}^{B}$ implicitly couples to $W_{\mathrm{nar}}$ through the query-key
mechanism: $\alpha_{ji}^{B,(h)} = f(Q_B(h + U_{ij}^{(A)}(h)), K_B(\cdot))$.

More precisely, writing the composed operation with attention weight dependence:
\[
    U_{ij}^{BA}(h) = \bar{\alpha}_{ji}^{B}\!\big(h + U_{ij}^{(A)}(h)\big)\,
    P_{W_{\mathrm{mor}}}\big(h + U_{ij}^{(A)}(h)\big),
\]
the cross-term acts as a map from $W_{\mathrm{nar}}$ into $W_{\mathrm{mor}}$ through the
dependence of layer $B$'s attention weights on layer $A$'s output. This is non-zero
whenever layer $B$ attends differently to the post-layer-$A$ residual than it would to
the pre-layer-$A$ residual.

For the reverse ordering:
\[
    U_{ij}^{AB}(h) = \bar{\alpha}_{ji}^{A}\!\big(h + U_{ij}^{(B)}(h)\big)\,
    P_{W_{\mathrm{nar}}}\big(h + U_{ij}^{(B)}(h)\big)
    + \text{layer-}B\text{ terms}.
\]
Since the attention weights of layer $A$ now depend on layer $B$'s output (which has a
$W_{\mathrm{mor}}$ component), the result differs from $U_{ij}^{BA}$. The non-commutativity
$U_{ij}^{AB} \neq U_{ij}^{BA}$ is the signature of dynamic non-commutativity from layer
composition (Remark~\ref{rem:non-abelian-source}).
\end{proof}

\begin{remark}[Source of Non-Commutativity]
\label{rem:non-abelian-source}
The structure group $G = \prod_{k \in K} O(d_k)$ is \emph{abelian}: each block $O(d_k)$
acts independently on $W_k$, and distinct blocks commute. The non-commutativity
$U_{ij}^{AB} \neq U_{ij}^{BA}$ therefore does \textbf{not} arise from the gauge group
itself. It arises from the \emph{dynamic dependence} of layer $l+1$'s attention weights on
layer $l$'s residual stream output---the fact that $\alpha^{(B)}$ is a function of the
post-layer-$A$ hidden state, not the pre-layer-$A$ state.

This is best understood by analogy with gauge theory coupled to matter fields. In
Yang--Mills theory with fermions, the gauge group may be abelian (as in $U(1)$
electrodynamics), yet the effective dynamics is non-trivial because the gauge field
couples to the matter field (the Dirac spinor), which in turn back-reacts on the gauge
field. In our setting, the residual stream plays the role of the matter field, and the
attention weights play the role of the gauge field: each layer's connection coefficients
depend on the current state of the residual stream, which was shaped by all preceding
layers. The non-commutativity is a property of this coupled dynamics, not of the
underlying symmetry group.
\end{remark}

% ------------------------------------------------------------------------------
\subsection{Gauge Invariance}
\label{sec:gauge}

A fundamental requirement of any gauge-theoretic framework is that physically meaningful
quantities are invariant under gauge transformations. We now verify that the block-specific
curvature norms satisfy this requirement.

\begin{proposition}[Gauge Invariance of Block-Specific Holonomy Norms, {\cite[Prop.~1]{LCESA}}]
\label{prop:gauge-invariance}
Under the gauge transformation $U_{ij} \mapsto g_j\, U_{ij}\, g_i^{-1}$ with $g_i, g_j
\in G = \prod_{k \in K} O(d_k)$, the block-specific holonomy norms are invariant:
\[
    \big\|[F^g_{ijk}]_k - \Pi_k\big\|_F = \big\|[F_{ijk}]_k - \Pi_k\big\|_F
    \qquad \forall\, k \in K,
\]
where $F^g_{ijk}$ is the gauge-transformed curvature and $[F^g_{ijk}]_k = \Pi_k\,
F^g_{ijk}\, \Pi_k$.
\end{proposition}

\begin{proof}
Since $g_i \in G = \prod_{k \in K} O(d_k)$, write $g_i = (g_i^{(\mathrm{min})}, \ldots,
g_i^{(\mathrm{pos})})$ with each $g_i^{(k)} \in O(d_k)$. Each $g_i$ acts on $\mathbb{R}^d$
by applying $g_i^{(k)}$ within $W_k$ and leaving $W_l$ ($l \neq k$) unchanged. Therefore
$g_i$ maps each $W_k$ to itself, and the oblique projectors satisfy $g_i\, \Pi_k\,
g_i^{-1} = \Pi_k$ (the subspaces are gauge-invariant).

The gauge-transformed holonomy deviation is $F^g_{ijk} = g_k\, F_{ijk}\, g_k^{-1}$. The
block-specific holonomy norm transforms as:
\begin{align*}
    [F^g_{ijk}]_k
    &= \Pi_k\, (g_k\, F_{ijk}\, g_k^{-1})\, \Pi_k \\
    &= g_k^{(k)}\, (\Pi_k\, F_{ijk}\, \Pi_k)\, (g_k^{(k)})^{-1}
       & \text{(since } g_k\, \Pi_k = \Pi_k\, g_k \text{ on } W_k\text{)} \\
    &= g_k^{(k)}\, [F_{ijk}]_k\, (g_k^{(k)})^{-1}.
\end{align*}
Therefore:
\begin{align*}
    \big\|[F^g_{ijk}]_k - \Pi_k\big\|_F
    &= \big\|g_k^{(k)}\, [F_{ijk}]_k\, (g_k^{(k)})^{-1} - \Pi_k\big\|_F \\
    &= \big\|g_k^{(k)}\, \big([F_{ijk}]_k - \Pi_k\big)\, (g_k^{(k)})^{-1}\big\|_F
       & \text{(since } g_k^{(k)}\, \Pi_k\, (g_k^{(k)})^{-1} = \Pi_k\text{)} \\
    &= \big\|[F_{ijk}]_k - \Pi_k\big\|_F,
\end{align*}
where the last step uses the Frobenius norm invariance under orthogonal conjugation:
$\|Q A Q^\top\|_F = \|A\|_F$ for all $Q \in O(n)$ and $A \in \operatorname{Mat}(n,
\mathbb{R})$, verified by $\|Q A Q^\top\|_F^2 = \operatorname{tr}(Q A^\top A Q^\top) =
\operatorname{tr}(A^\top A) = \|A\|_F^2$.

Note: this proof does \emph{not} require the subspaces to be orthogonal or the holonomy deviation
to be block-diagonal. The gauge invariance of block-specific holonomy norms holds for
any independent subspaces with oblique projectors.
\end{proof}

\begin{remark}[Physical Meaning]
The gauge invariance established in Proposition~\ref{prop:gauge-invariance} has a direct
operational interpretation: the diagnosis of \emph{which} standpoint layer failed (i.e.,
which $\|F_{ijk}\|_k$ is large) is a geometric invariant of the fiber bundle, not an
artifact of the choice of basis within each subspace. Different choices of orthonormal
basis for $W_k$---corresponding to different gauge choices---yield the same curvature
norms. This ensures that the diagnostic framework of~\cite{LCESA} (classifying failures
by which layer's curvature is elevated) is well-defined.
\end{remark}

% ------------------------------------------------------------------------------
\subsection{Connection to the Commitment Architecture}
\label{sec:correspondence}

We conclude by making explicit the correspondence between the present geometric
framework, the commitment architecture of~\cite{self-jurisdiction}, and the abstract
low-curvature standpoint theory of~\cite{LCESA}. This table demonstrates that the
three formulations---logical, geometric, and architectural---are three faces of a single
mathematical structure.

\begin{table}[h]
\centering
\caption{Correspondence between the commitment architecture~\cite{self-jurisdiction},
the geometric framework~\cite{LCESA}, and the transformer instantiation
(Section~\ref{sec:instantiation}).}
\label{tab:correspondence}
\begin{tabular}{@{}p{3.8cm}p{4.2cm}p{5.2cm}@{}}
\toprule
\textbf{Commitment Arch.} & \textbf{Geometry (LCESA)} & \textbf{Transformer Instantiation} \\
\midrule
$M_t$: self-jurisdiction state &
$\Psi_{x_t}$: standpoint section &
$\Pi_W h_{x_t}$: oblique-projected residual stream \\[4pt]
$\mathrm{Bind}(c_t, a_t, S_t, \ell_t)$ &
Connection transport $\Pi_\Psi(j \leftarrow i)$ &
Attention operation $U_{ij}$ \\[4pt]
Warrant Conservation &
Flat transport condition (1) &
Attention preserves standpoint norm \\[4pt]
Self-Binding &
Connection $A_\Psi$ &
Residual stream accumulation $h + \sum_l \mathrm{attn}^{(l)}(h)$ \\[4pt]
Repair Obligation &
Flat transport condition (3) &
Curvature $\|F_{ijk}\|_k > \epsilon \to$ repair trigger \\[4pt]
Binding failure &
Non-zero curvature $F_{ijk}$ &
Path-dependence of attention transport \\[4pt]
Layer $k$ binding failure &
$\|F_{ijk}\|_k > 0$ &
$\|[F_{ijk}]_k - \Pi_k\|_F > 0$ \\[4pt]
12 diagnostic modes &
Curvature decomposition by layer and scenario type &
Block-specific curvature norms across T1/T2/T3 conversations \\
\bottomrule
\end{tabular}
\end{table}

\begin{theorem}[Holonomy Deviation as Binding Failure: Formal Equivalence]
\label{thm:curvature-binding}
Let $\Psi_{x_i}$ be the standpoint state at event $x_i$, and let $U_{ij} \cdot U_{jk}$ be
the composite transport through event $x_j$ to event $x_k$. Then for each layer $k \in K$:
\[
    \|F_{ijk}\|_k > 0 \quad\Longleftrightarrow\quad \text{the commitment encoded in
    layer $k$ was not preserved through event $x_j$.}
\]
\end{theorem}

\begin{proof}
This is a restatement of Theorem~\ref{thm:binding-failure} in the language
of~\cite{self-jurisdiction}. The commitment architecture defines binding failure as the
situation in which the self-jurisdiction state $M_t$ at time $t$ does not cohere with the
prior committed state. In the geometric instantiation, $M_t$ corresponds to the standpoint
state $\Psi_{x_t}$ (Table~\ref{tab:correspondence}), and the commitment is preserved if
and only if parallel transport from $x_i$ to $x_k$ (through the pressure event $x_j$)
yields the same result as the expected direct transport. The curvature $F_{ijk}$ measures
exactly this discrepancy: $F_{ijk} = I_d$ means transport is path-independent (commitment
preserved), and $[F_{ijk}]_k \neq \Pi_k$ means layer $k$'s commitment was violated.

For the forward direction: if $\|F_{ijk}\|_k > 0$, then $[F_{ijk}]_k \neq \Pi_k$,
so the $k$-th block of the composite transport $U_{ij} \cdot U_{jk}$ differs from the
$k$-th block of $U_{ik}^{\exp}$. The event $x_j$ induced a change in $\psi_k$ that was
not present in the expected (flat) transport. By the correspondence $U_{ij} \leftrightarrow
\mathrm{Bind}$ (Table~\ref{tab:correspondence}), this change is a binding failure: the
commitment encoded in $W_k$ was not conserved.

For the converse: if the commitment in layer $k$ was not preserved, then $\psi_k(x_k)
\neq \psi_k^{\exp}(x_k)$ (the standpoint component at $x_k$ differs from the expected
value). Since $\psi_k(x_k) = [U_{ij} \cdot U_{jk}]_k\, \psi_k(x_i)$ and
$\psi_k^{\exp}(x_k) = [U_{ik}^{\exp}]_k\, \psi_k(x_i)$, we have $[U_{ij} \cdot U_{jk}]_k
\neq [U_{ik}^{\exp}]_k$, hence $[F_{ijk}]_k \neq \Pi_k$ and $\|F_{ijk}\|_k > 0$.
\end{proof}

\begin{remark}[The Low-Curvature Attractor]
\label{rem:low-curvature-attractor}
The central thesis of~\cite{LCESA} is that functional selfhood in a conversational agent
corresponds to the \emph{low-curvature region} of the standpoint bundle:
\[
    \mathcal{A}_{\mathrm{self}} = \{\Psi \in \Gamma(E) : \|F_\Psi\| < \epsilon\},
\]
where $\Gamma(E)$ denotes the space of sections (standpoint states across all events) and
$\epsilon > 0$ is a threshold. Theorem~\ref{thm:curvature-binding} shows that membership
in $\mathcal{A}_{\mathrm{self}}$ is equivalent to the absence of binding failure across all
standpoint layers: the agent's commitments are preserved through all consistency
obligations. The present instantiation (Section~\ref{sec:instantiation}) makes this
region computable: $\|F_\Psi\|$ is the Frobenius norm of the holonomy deviation tensor, which can
be estimated from attention patterns and value matrices of any deployed transformer.
\end{remark}
